---
title: "TargetRepositioningProstateCancer"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{TargetRepositioningProstateCancer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction
***DiscoNet*** is an R package containing an ensemble of functions to automatically extract node features from complex biological network that can later be used by machine learning algorithms for prediction, classification, etc.
The main categories of features that can be extracted from a set of nodes (`start_nodes`) using  ***DiscoNet*** are: 
 - propagation-based features: simple one-to-one distances between `start_nodes` and other nodes in the network, calculated through either shortest paths or random walks. The distances can be directed, inversely directed, or undirected. (extract_by_shp, extract_by_shp_inv, extract_by_shp_ind, extract_by_rwr,extract_by_rwr_inv, extract_by_rwr_ind).
 - topological metrics and similarities: common metrics such as degree, betweenness, or centrality are extracted, as well as node similarity based on node neighborhoods (extract_topolo).
 - module-based: Based on clusters previously calculated form the network (else it will compute clusters based on already available methods from the igraph R package), will detect if some `start_nodes` are within these clusters and their proximity to each cluster based on the available distance measures (available within propagation-based features) (extract_cluster). 
 - signature-based: Based on a user-own disease signatures (list of genes or proteins), different proximity between `start_nodes` and these genes can be calculated (In signature, percentage of neighboors in signatures, distance to the whole signature, and distance to subsets of signatures based on clusters separation) (extract_by_sig).

This vignette shows a script example on how to use DiscoNet to extract features from the A3LNetwork to do therapeutic target repositioning on prostate cancer (supervised classification).

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r load library}
# You'll need these four libraries to run this vignette, they can be be easily installed like so:
# if (!require("BiocManager", quietly = TRUE))
#   install.packages("BiocManager")
# 
# BiocManager::install("devtools")
# BiocManager::install("tidyverse")
# BiocManager::install("igraph")
# devtools::install_github("https://github.com/MilanPicard/DiscoNet")

library(DiscoNet)
library(tidyverse)
library(igraph)
```

# Feature extraction
***Disconet*** needs at least a biological network from which to extract features. If no `start_nodes` are given (which are the nodes for which to extract features), features will be extracted for every node in the network.
In this package, a three layer network called A3LNetwork (protein+gene+go) of medium size is preloaded. It is comprised of a PPI network (from stringdb), a gene layer connected in a directed manner to the PPI layer, and a GO layer connected to the gene layer. How the network was built can be accessed in data-raw/DATASET.R
The type of nodes are described as either a prefix in the node name ('Prot_', 'Gene_', 'GO_'), or as node attribute (`type`)
```{r load necessary data}
A3LNetwork = DiscoNet::A3LNetwork
summary(A3LNetwork)
table(V(A3LNetwork)$type)
# IGRAPH 5f51ddf DN-B 7538 53919 -- 
# + attr: name (v/c), type (v/c)
# 
#    gene      go protein 
#    3757    1916    1865 
```

For target repositioning against prostate cancer, we are interested in extracting features for every protein in the network, as any of them could be of interest.
Let's calculate some basic features like topological metrics or simple distances to other nodes in the network
```{r Extracting basic features for each protein}
# Get all proteins beginning with Prot_
all_proteins = vnames(A3LNetwork, pattern = "Prot_")

# Calculate topological features for the 1 865 proteins in the network
TopologicalMetrics = extract_topolo(A3LNetwork, all_proteins)
dim(TopologicalMetrics)

# Calculate for the 1 865 proteins their random walk distance to every other nodes
# This will take a bit of time, especially without nCores = 1 (no parallel computing)
# Faster distance calculation include the shortest path ones.
# shp_dist = extract_by_shp(Graph = A3LNetwork, start_nodes = all_proteins)
rwr_dist = extract_by_rwr(Graph = A3LNetwork, start_nodes = all_proteins, nCores = 1)
dim(rwr_dist)
# Additional distance measures can be obtained with extract_by_rwr_inv or extract_by_rwr_ind
```

More complex features can also be extracted when exploiting community detection on the network, including module-based features as well as signature-based features.
```{r Extracting module-based features for each protein}
# Based on precalculated clusters done on A3LNetwork
Protein_clusters = DiscoNet::data_prostate_cancer$Clusters

# They can also be calculated using the following code
# # Create PPI subgraph
# A3LNetwork_prot =igraph::induced_subgraph(A3LNetwork, vids = all_proteins)
# # Calculate clusters, optionnaly, you don't have to calculate clusters if you have precalculated modules you want to use.
# A3LNetwork_prot_clust = extract_cluster(Graph = A3LNetwork_prot, start_nodes = all_proteins, only_cluster = TRUE)
# # Only take clusters of node size higher than 10 
# A3LNetwork_prot_clust = A3LNetwork_prot_clust[nbrs(A3LNetwork_prot_clust) > 10] # Results in 55 modules


# Proximity of every  protein to each cluster is calculated based on the distance matrix that we just calculated. If no dist_matrix is given, extract_cluster will internally run extract_by_shp to get distances
# If no clusters are given, it will calculate its own based on the whole network using cluster_walktrap and cluster_louvain from the igraph R package.
Dist_cluster = extract_cluster(A3LNetwork, all_proteins, dist_matrix = rwr_dist, Cluster_list = Protein_clusters)
dim(Dist_cluster)
```

Signature-based features necessitates a list of genes (or proteins) to be used as a signature that the user must have. For this specific task, a preloaded signature is included within ***DiscoNet***
```{r Extracting signature-based features for each protein}
data_prostate_cancer = DiscoNet::data_prostate_cancer

# Clusters should be converted to the same ID than the gene layer ID
Gene_clusters = data_prostate_cancer$Clusters_gene
# Remove genes not present in the network, or else an error will be returned
Gene_clusters = lapply(Gene_clusters, function(x) intersect(x, vnames(A3LNetwork)))

# Get preloaded gene signatures for prostate cancer, it contains a list of signatures and also known therapeutic targets.

# Remove genes not present in the network (it is rather a small network), resulting in 25 mutated genes and 5 differentially expressed genes.
Signature_for_prostate_cancer = lapply(data_prostate_cancer$Signatures, function(x) intersect(x, vnames(A3LNetwork)))

# Extract signature-based features for 1 865 proteins
Dist_signature = extract_by_sig(A3LNetwork, all_proteins, Signature_for_prostate_cancer, Gene_clusters, rwr_dist)
Dist_signature %>% dim
```

# Feature selection
Once some features are extracted, it is often necessary to reduce their dimensionality before applying machine learning algorithm
***DiscoNet*** make accessible to supervised variable selection method, a quick but robust selection based on information gain with bootstraping, or a longer but more powerful selection based on Adaptive LASSO.
```{r Variable selection}
# merge all feature datasets together. 
Full_features = purrr::reduce(.x = list(TopologicalMetrics, rwr_dist, Dist_cluster, Dist_signature), 
                              merge, by = c('Target'), all = T)

Full_features %>% dim
# That is a lot of features, so variable selection will be done in order to keep only the most informative ones.
# To run information gain variable selection, a class must be given.
# In this vignette, the class will be based on whether or not a protein is a known therapeutic target for prostate cancer
# data_prostate_cancer$Targets contains 13 known targets. The other proteins are therefore assumed negatives.
class = Full_features$Target %in% data_prostate_cancer$Targets

# Create a training set and a test set
# training set contains all positive observations and a random selection of 200 negative observations on which to train
set.seed(123)
index = c(which(class),sample(which(!class), 300))

train.x = Full_features[index, -1]
train.y = class[index]

test.x = Full_features[-index, -1]
test.y = class[-index]

# Run information gain on different bootstraps
infgain_full = InformationGain_Bootstrap(df = train.x, class = train.y, nbr_of_boot = 3, seed = 123)
# Take the best 100 features (with the highest information gain)
best_features = infgain_full$feature[1:100]
```

# Machine learning classification
```{r Prediction using KKNN}
# Using your favorite machine learning algorithms, you can now use these features to predict new therapeutic targets for prostate cancer
# Example:
# if (!require("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")
# 
# if (!require("kknn", quietly = TRUE))
#     BiocManager::install("kknn")
# 
# library("kknn")
# 
# # Using the k-nearest neighbors algorithm to train of these features on the training set.
# set.seed(123)
# KNN_mod = kknn::train.kknn(Y~., data = data.frame(Y = as.factor(train.y), train.x[, best_features]),  ks = 5,  distance = 2)
# conf_mat = table(preds = KNN_mod$fitted.values[[1]], actuals = KNN_mod$data$Y)
# print(paste0("Accuracy: ", round(sum(diag(conf_mat))/sum(conf_mat), 2)))
# print(paste0("Sensibility: ", round(conf_mat[1, 1]/sum(conf_mat[1, ]), 2)))
# print(paste0("Sensitivity: ", round(conf_mat[2, 2]/sum(conf_mat[2, ]), 2)))
# 
# # Predict new potential targets on the bigger test set.
# PotentialNewTarget = Full_features$Target[-index][predict(KNN_mod, test.x[, best_features]) == TRUE]

```









